### 2024-11-26 paper/paper.qmd
Data section uploaded
Data section uploaded

---
title: "Analysis and Prediction of Shared Bicycle Usage at the University of Toronto St. George Campus"
subtitle: "Main Finding TBD"
author: 
  - Haowei Fan
thanks: "Code and data are available at: https://github.com/HaoweiFan0912/Bikeshare-Forecast.git"
date: today
date-format: long
abstract: "Students and staff at the University of Toronto's St. George campus often face difficulties finding a bike-sharing parking spot when arriving at campus, and finding a bike when leaving, which highlights the need for effective prediction of future bike-sharing demand to support campus commuting infrastructure. This study uses Bayesian Poisson regression to predict the usage of bike-sharing stations at 27 locations on campus for specific years, months, and  four-hour intervals of the day. The results show that bike-sharing usage on campus has been increasing annually, with significant seasonal variations and peak usage between 8 am and 8 pm."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

This study found that although the usage at all stations within the University of Toronto campus has increased rapidly each year, there are significant seasonal differences. During winter, usage at all stations, regardless of the year, remains close to zero. Additionally, from January to September each year, the usage shows an upward trend, which then gradually declines over the following months. There are also significant differences in usage across different times of the day. The period from 8 am to 8 pm accounts for an average of 79.8% of daily usage, while the period from midnight to 8 am accounts for only 6.8%. Notably, the number of bikes checked out and parked at the 27 stations is generally balanced; however, there are four stations where there is a substantial discrepancy between the number of check-outs and returns.

@@

```{r}
#| label: fig-daily_usage
#| fig-cap: Daily usage from January 2017 to September 2024
#| echo: false
#| message: false
start_asdate <- start 
start_asdate$time <- as.Date(start_asdate$time)
daily_counts <- start_asdate %>%
  group_by(time) %>%
  summarise(total_count = sum(count))
ggplot(data = daily_counts, aes(x = time, y = total_count)) +
  geom_point() +
  labs(x = "Time", y = "Count") +
  theme_minimal()+  
  geom_smooth(method = "lm")  
```

>>>>

---
title: "Analysis and Prediction of Shared Bicycle Usage at the University of Toronto St. George Campus"
subtitle: "Main Finding TBD"
author: 
  - Haowei Fan
thanks: "Code and data are available at: https://github.com/HaoweiFan0912/Bikeshare-Forecast.git"
date: today
date-format: long
abstract: "Students and staff at the University of Toronto's St. George campus often face difficulties finding a bike-sharing parking spot when arriving at campus, and finding a bike when leaving, which highlights the need for effective prediction of future bike-sharing demand to support campus commuting infrastructure. This study uses Bayesian Poisson regression to predict the usage of bike-sharing stations at 27 locations on campus for specific years, months, and  four-hour intervals of the day. The results indicate that, on September 26, 2025, from 4:00 to 8:00 AM, usage at 8 stations will increase by 3, 14 stations by 2, 3 stations by 1, 1 station remains unchanged, and 1 station will decrease by 1 compared to the same period in 2024."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

This study found that although the usage at all stations within the University of Toronto campus has increased rapidly each year, there are significant seasonal differences. During winter, usage at all stations, regardless of the year, remains close to zero. Additionally, from January to September each year, the usage shows an upward trend, which then gradually declines over the following months. There are also significant differences in usage across different times of the day. The period from 8 am to 8 pm accounts for an average of 79.8% of daily usage, while the period from midnight to 8 am accounts for only 6.8%. It is noteworthy that this project forecasts an average increase of 2 uses across 27 stations between 4:00 AM and 8:00 AM on September 26, 2025, compared to the same time period in 2024. Specifically, usage at 8 stations is expected to increase by 3, 14 stations by 2, and 3 stations by 1, while 1 station will remain unchanged and another will decrease by 1.

@@

@fig-daily_usage shows the daily usage totals of 27 shared bicycle stations from January 1, 2017, to September 30, 2024. It can be observed that the overall usage exhibits a significant upward trend, particularly after 2021, where usage fluctuations increased noticeably. In 2024, several peaks in daily usage reached historical highs, indicating a substantial growth in user demand. Additionally, the data displays some seasonal variations, with noticeable declines during the winter months and increases during spring and summer.
```{r}
#| label: fig-daily_usage
#| fig-cap: Daily usage from January 2017 to September 2024
#| echo: false
#| message: false
start_asdate <- start 
start_asdate$time <- as.Date(start_asdate$time)
daily_counts <- start_asdate %>%
  group_by(time) %>%
  summarise(total_count = sum(count))
ggplot(data = daily_counts, aes(x = time, y = total_count)) +
  geom_point() +
  labs(x = "Time", y = "Count") +
  theme_minimal()+  
  geom_smooth(method = "lm")  
```

hows the total usage and percentage of shared bicycles during different time intervals in a day from January 1, 2017, to September 30, 2024. It can be observed that the peak usage occurs between 12:00 to 16:00 (28.5%) and 16:00 to 20:00 (31.4%). The usage between 08:00 to 12:00 accounts for 19.9%, while the usage from 20:00 to 00:00 accounts for 13.4%. The time intervals with the lowest usage are 00:00 to 04:00 (3.6%) and 04:00 to 08:00 (3.2%).

```{r}
#| label: fig-timely_usage
#| fig-cap: Total usage during different time periods from January 2017 to September 2024
#| echo: false
#| message: false
options(scipen = 999)

start_astime <- start 
start_astime$time <- as.POSIXct(start_astime$time)

start_astime <- start_astime %>%
  mutate(time_interval = case_when(
    hour(time) >= 0 & hour(time) < 4 ~ "00:00 - 04:00",
    hour(time) >= 4 & hour(time) < 8 ~ "04:00 - 08:00",
    hour(time) >= 8 & hour(time) < 12 ~ "08:00 - 12:00",
    hour(time) >= 12 & hour(time) < 16 ~ "12:00 - 16:00",
    hour(time) >= 16 & hour(time) < 20 ~ "16:00 - 20:00",
    hour(time) >= 20 & hour(time) < 24 ~ "20:00 - 00:00"
  ))

total_counts <- start_astime %>%
  group_by(time_interval) %>%
  summarise(count = sum(count, na.rm = TRUE)) %>%
  mutate(percentage = count / sum(count) * 100)

ggplot(data = total_counts, aes(x = time_interval, y = count, fill = time_interval)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), 
            vjust = -0.5, size = 3) + 
  labs(x = "Time interval", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")
```

@fig-difference_usage shows the result of the average daily parking volume minus the departure volume at each station. It can be observed that most stations have balanced parking and departure volumes, but some stations show significant supply-demand differences. Specifically, stations such as Bay St / Charles St W – SMART and Bay St / Wellesley St W have noticeably higher departure volumes than parking, while College St / Huron St and College St / Henry St have higher parking volumes than departure, indicating an imbalance in supply and demand at these locations.

```{r, fig.width=8, fig.height=5}
#| label: fig-difference_usage
#| fig-cap: Difference of daily average of bicycle parking and departure volumes for each station
#| echo: false
#| message: false
options(scipen = 999)

# Create a new column for the time interval (every 1day)
start_1day <- start %>%
  mutate(date = as.Date(time))
stop_1day <- stop %>%
  mutate(date = as.Date(time))
start_1day <- start_1day %>% select(-time)
stop_1day <- stop_1day %>% select(-time)

# Group by from(to)_station_name and the new interval, then count the occurrences
start_1day1 <- start_1day %>%
  group_by(station_name, date) %>%
  summarise(total_count = sum(count), .groups = "drop") %>%
  ungroup()
stop_1day1 <- stop_1day %>%
  group_by(station_name, date) %>%
  summarise(total_count = sum(count), .groups = "drop") %>%
  ungroup()

average_counts_start <- start_1day1 %>%
  group_by(station_name) %>%
  summarize(average_count = mean(total_count, na.rm = TRUE))

average_counts_stop <- stop_1day1 %>%
  group_by(station_name) %>%
  summarize(average_count = mean(total_count, na.rm = TRUE))

merged_data <- merge(average_counts_start, average_counts_stop, by.x = "station_name", by.y = "station_name", all = TRUE)

merged_data <- merged_data %>%
  mutate(count_difference = ifelse((average_count.y - average_count.x) >= 0, 
                                   ceiling(average_count.y - average_count.x), 
                                   floor(average_count.y - average_count.x)))

ggplot(merged_data, aes(x = station_name, y = count_difference)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(
       x = "Station Name",
       y = "Parking volume minus departure volume")+
  ylim(-4, 3)+
  coord_flip()
```

### 2024-11-26 scripts/05-exploratory_data_analysis.qmd
Data section uploaded
Data section uploaded

```{r}
md <- readRDS(here::here("models/Ursula_Franklin_St___Huron_St_-_SMART.rds"))
new_data <- start %>% filter(station_name == "Ursula Franklin St / Huron St - SMART")
train <- new_data %>%
    mutate(
      hour = hour(time),
      day = day(time),
      month = month(time),
      year = year(time)
    )
new_data$Predicted <- predict(md, newdata = train)
new_data <- new_data[new_data$count <= 8, ]
ggplot(new_data, aes(x = count, y = exp(Predicted))) +
  geom_point(color = "blue", alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") + 
  labs(
    title = "Scatter Plot of True vs Predicted Values",
    x = "True Values",
    y = "Predicted Values"
  ) +
  theme_minimal()
```

>>>>

```{r}
md <- readRDS(here::here("models/Ursula_Franklin_St___Huron_St_-_SMART.rds"))
new_data <- start %>% filter(station_name == "Ursula Franklin St / Huron St - SMART")
train <- new_data %>%
    mutate(
      hour = hour(time),
      day = day(time),
      month = month(time),
      year = year(time)
    )
new_data$Predicted <- predict(md, newdata = train)
new_data <- new_data[new_data$count <= 8, ]
ggplot(new_data, aes(x = count, y = exp(Predicted))) +
  geom_point(color = "blue", alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") + 
  labs(
    title = "Scatter Plot of True vs Predicted Values",
    x = "True Values",
    y = "Predicted Values"
  ) +
  theme_minimal()
```
```{r}
ut_stations <- c("Madison Ave / Bloor St W", "Bloor St W / Huron St", "St. George St / Bloor St W", 
                 "Sussex Ave / St George St", "Spadina Ave / Sussex Ave", "Spadina Ave / Harbord St - SMART",
                 "St. George St / Hoskin Ave", "Spadina Ave / Willcocks St", "St. George St / Willcocks St",
                 "Willcocks St / St. George St", "Queen's Park / Bloor St W", "Queen's Park Cres W / Hoskin Ave",
                 "Wellesley St W / Queen's Park Cres", "Queen's Park Cres E / Grosvenor St - SMART", 
                 "Bay St / Bloor St W (East Side)", "Bay St / Bloor St W (West Side)", "Bay St / Charles St W - SMART",
                 "St. Joseph St / Bay St - SMART", "Bay St / St. Joseph St", "Bay St / Wellesley St W", 
                 "Ursula Franklin St / Huron St - SMART", "Ursula Franklin St / St. George St - SMART", "Galbraith Rd / King's College Rd",
                 "College St / Huron St", "College St / Henry St ", "Queens Park Cres / College St ", "University Ave / College St (East)")

result <- data_frame(station_name = ut_stations, time = as.POSIXct("2025-09-26 04:00:00", format = "%Y-%m-%d %H:%M:%S"))

for (station in ut_stations) {
  station_data <- subset(start, station_name == station)
  path <- paste(paste0("models/", gsub("[ /]", "_", station), ".rds"))
  md <- readRDS(here::here(path))
  target <- data_frame(
    hour = 4,
    day = 26,
    month = 9,
    year = 2026
  )
  target$predicted <- ceiling(exp(predict(md, newdata = target)))
  target$station_name <- station
  target <- target %>% select(-hour,-day,-month,-year)
  result[result$station_name == station, "prediction"] <- target$predicted
}

data2024 <- start
for (station in ut_stations) {
  date <- "2024-09-26 04:00:00"
  exists <- nrow(data2024[data2024$station_name == station & data2024$time == date, ]) > 0

  if (!exists) {
    new_row <- data.frame(
      station_name = station,
      time = date,
      count = 0
    )
    data2024 <- rbind(data2024, new_row)
  }
}

data2024 <- data2024[data2024$time == "2024-09-26 04:00:00", ]

names(result)[names(result) == "time"] <- "target_time"

data_fi <- merge(data2024, result, by = "station_name")

data_fi$growth <- data_fi$prediction - data_fi$count

data_fi

n <- sum(data_fi$growth)/27
n
```

### 2024-11-26 scripts/08-model_validation.qmd
Data section uploaded
Data section uploaded

```{r}
poisson_glm <- readRDS(here::here("models/Galbraith_Rd___King's_College_Rd.rds"))

data <- read_parquet(here::here("data/02-analysis_data/start.parquet")) %>% filter(station_name == "Galbraith Rd / King's College Rd
")%>%
    mutate(
      hour = hour(time),
      day = day(time),
      month = month(time),
      year = year(time)
    )

@@

```{r}
# Load necessary library
library(bayesplot)
model=poisson_glm
# 1. Trace Plots
mcmc_trace(model, pars = c("(Intercept)", "your_predictor_names_here"))

# 2. Posterior Distribution
mcmc_dens_overlay(model, pars = c("(Intercept)", "your_predictor_names_here"))

# 3. R-hat values to check convergence
summary(model)$rhat

# 4. Effective sample size
summary(model)$n_eff

# 5. Posterior Predictive Checks
pp_check(model, plotfun = "dens_overlay", nreps = 100) # Overlay of predictive densities

# 6. Residual Analysis
residuals <- residuals(model, type = "response")
fitted <- fitted(model)
plot(fitted, residuals, main = "Residuals vs Fitted", xlab = "Fitted values", ylab = "Residuals")
abline(h = 0, col = "red")

posterior_draws <- as.matrix(model)

# Compute credible intervals
credible_intervals <- apply(posterior_draws, 2, function(x) quantile(x, probs = c(0.025, 0.975)))
print(credible_intervals)

# Replace "your_predictor_names_here" with the actual names of your predictors

>>>>

```{r}
#| echo: false
#| include: false
#| warning: false
#| message: false
#### Workspace setup ####
set.seed(912)
# List of necessary packages
packages <- c("arrow", "lubridate", "dplyr", "brms", "rstanarm","tidyverse","forecast","tseries","bayesplot")
# Install missing packages
missing_packages <- packages[!(packages %in% installed.packages()[, "Package"])]
install.packages(missing_packages)
# Load packages
lapply(packages, library, character.only = TRUE)

#### Loading models ####
poisson_glm <- readRDS(here::here("models/Galbraith_Rd___King's_College_Rd.rds"))
```

```{r}
data <- read_parquet(here::here("data/02-analysis_data/start.parquet")) %>% filter(station_name == "Galbraith Rd / King's College Rd
")%>%
    mutate(
      hour = hour(time),
      day = day(time),
      month = month(time),
      year = year(time)
    )

@@

```{r}
# 5. Posterior Predictive Checks
pp_check(model, plotfun = "dens_overlay", nreps = 100) # Overlay of predictive densities
```

```{r}
# 6. Residual Analysis
residuals <- residuals(model, type = "response")
fitted <- fitted(model)
plot(fitted, residuals, main = "Residuals vs Fitted", xlab = "Fitted values", ylab = "Residuals")
abline(h = 0, col = "red")
```

```{r}
posterior_draws <- as.matrix(model)

# Compute credible intervals
credible_intervals <- apply(posterior_draws, 2, function(x) quantile(x, probs = c(0.025, 0.975)))
print(credible_intervals)
```
